### 2023-01-09 paper-thesis/chapters/11-variable-capture.tex
Work on writing

TODO Add reference to types and programming languages

The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture. This section will explore several ways of solving this. An example is the following: What is the type of this expression (a polymorphic identity function)?
\begin{lstlisting}
\T : Type. \T : T. T
\end{lstlisting}

The algorithm so far would tell you it is \verb|T : Type -> T : T -> T|. Given the scoping rules of the language, that is equivalent to \verb|T : Type -> x : T -> x|. However, the correct answer would be \verb|T : Type -> x : T -> T|. There is no way of expressing this type without renaming a variable.

>>>>

The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture. Variable capture is the phenomenon of free variables in a term becoming bound when a naive substitution happens \cite{tapl}. This section will explore several ways of solving this.

An example term where this problem occurs is the following: What is the type of this expression (a polymorphic identity function)?
\begin{lstlisting}
\T : Type. \T : T. T
\end{lstlisting}
The algorithm so far would tell you it is \verb|T : Type -> T : T -> T|. Given the scoping rules of the language, that is equivalent to \verb|T : Type -> x : T -> x|. However, the correct answer would be \verb|T : Type -> x : T -> T|. There is no way of expressing this type without renaming a variable.

### 2023-01-09 paper-thesis/chapters/20-adding-booleans.tex
Work on writing

\begin{enumerate}
	\item Define the parsing rules for the new feature
	\item Create a new file, \verb|tp_bools.stx| and import this file in the main \verb|type_check.stx| file.
	\item In this file, define a case for the \verb|betaReduceHead| rule for the constructors that were added. Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary.
	\item In this file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
\end{enumerate}

>>>>

\begin{enumerate}
	\item Define the parsing rules for the new feature
	\item Create a new file, \verb|tp_bools.stx| and import this file in the main \verb|type_check.stx| file.
	\item In the new file, define a case for the \verb|betaReduceHead| rule for the constructors that were added. Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary. It is necessary iff the new constructors are not always eliminated by \verb|betaReduceHead|.
	\item In the new file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
\end{enumerate}

### 2023-01-09 paper-thesis/chapters/21-adding-inference.tex
Work on writing

There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others. However, we would like to avoid implementing an algorithm at all, instead using Statix' built-in first-order unification to do the type inference for us. Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code (since Statix is not a general-purpose programming language), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.

The inference that we will build is a very simple unification system, where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the free variable to be equal to the other variable. This process is called \emph{first-order unification}.

There are some situations in which this approach fails, but in most real-world scenarios it works perfectly. For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:

\begin{lstlisting}
let f = _;
\x : Type.
\g: (_: (f x) -> Bool).
g true
\end{lstlisting}
We know that \verb|f| is a function from \verb|Type -> Type|, but it fails to infer the value of \verb|f|. Because of the way that \verb|g| is used, the type checker asserts that $\beq{f x}{Bool}$. Since x is declared as a function argument and it is completely free, this means that for any \verb|x|, \verb|f x = Bool|. But the rule above is not powerful enough to derive all of this, so it fails.

\section{Using Statix' first-order unification for inference}
\label{statix-inference}

First, we introduce a new constructor, \verb|Infer : Expr -> Expr|. This will be used to store metavariables which we want to infer. This is a hack we use, ideally we would like to query Statix to see if a metavariable is initialized, but as discussed in section \ref{strength-inference} this is not possible.

The constructor is introduced when we encounter a \verb|_| variable, a sign that something needs to be inferred. When type-checking this, we create an infer constructor. The constraint at the bottom cannot be solved immediately, since the \verb|q| variable is unbound, but if \verb|q| is bound later the constraint will be evaluated.

\begin{lstlisting}
typeOfExpr_(s, Var(Syn("_"))) = (Infer(q), qt) :-
	(_, qt) == typeOfExpr(sEmpty(), q).
\end{lstlisting}
When an \verb|Infer| needs to be type-checked, the logic is very similar, except that we don't need to generate a new metavariable:

@@

Finally, there is a case for if we encounter two \verb|Infer|s. Ideally, we would look at whether one infer is instantiated, and apply the rules above. However, this is not possible since we cannot do such queries in Statix. Instead, we say that they must be exactly equal and we hope for the best. There are situations where this rule fails (where e1 and e2 are beta-equal but not identical), but this works well enough for practical use.
\begin{lstlisting}
expectBetaEq_((_, Infer(e1)), (_, Infer(e2))) :-
    e1 == e2.
\end{lstlisting}

>>>>

There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others. One algorithm for unification is \emph{first-order unification}, where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable. There are some situations in which this approach fails, but in most real-world scenarios it works perfectly. For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:

\begin{lstlisting}
let f = _;
\x : Type.
\g: (_: (f x) -> Bool).
g true
\end{lstlisting}

We know that \verb|f| is a function from \verb|Type -> Type|, but it fails to infer the value of \verb|f|. Because of the way that \verb|g| is used, the type checker asserts that $\beq{f x}{Bool}$. Since x is declared as a function argument and it is completely free, this means that for any \verb|x|, \verb|f x = Bool|. But the rule above is not powerful enough to derive this, so it fails.

\section{Inference in Statix}
\label{statix-inference}

We would like to avoid implementing an algorithm at all, instead using Statix' built-in first-order unification to do the type inference for us. Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code (since Statix is not a general-purpose programming language), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.

However, we cannot immediately use Statix' built-in first-order unification (which acts in the meta language, Statix) to implement first-order unification in the object language. Ideally when implementing beta equality we would match on $\beq{e_1}{e_2}$ where $e_1$ is a free variable, but Statix does not allow for querying whether variables are free.

Instead, we will be implementing a novel, less powerful form of first-order unification. This will work by explicitly denoting which variables \emph{could be} free, and explicitly handling these cases in a way that approximates first-order unification. We will denote this algorithm as \emph{approximated first-order unification (AFOU)}.

\section{Implementing AFOU}
\label{implementing-inference}

First, we introduce a new constructor \verb|Infer : Expr -> Expr|, which denotes the variables which could be free. The constructor is introduced when we encounter a \verb|_| variable, a marker that something needs to be inferred.

\begin{lstlisting}
typeOfExpr(s, Var(Syn("_"))) = (Infer(q), qt) :-
	(_, qt) == typeOfExpr(sEmpty(), q).
\end{lstlisting}

Note that the type of \verb|typeOfExpr| has changed, it is now

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr * Expr
\end{lstlisting}

\section{Analysis of the power of AFOU}
\label{analysing-inference}

\begin{comment}

\section{Using Statix' first-order unification for inference}

When an \verb|Infer| needs to be type-checked, the logic is very similar, except that we don't need to generate a new metavariable:

@@

Finally, there is a case for if we encounter two \verb|Infer|s. Ideally, we would look at whether one infer is instantiated, and apply the rules above. However, this is not possible since we cannot do such queries in Statix. Instead, we say that they must be exactly equal and we hope for the best. There are situations where this rule fails (where e1 and e2 are beta-equal but not identical), but this works well enough for practical use.
\begin{lstlisting}
expectBetaEq_((_, Infer(e1)), (_, Infer(e2))) :-
    e1 == e2.
\end{lstlisting}
\end{comment}
